Investigating magnetic dynamo evolution with TESS field dwarfs

### 2017-10-03 ms.tex
still ridiculously long, this isn't funny

Angular momentum is carried away though magnetically driven stellar winds,
which slows the star’s rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star’s surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
While rotation periods have previously been measured from starspot-induced
flux modulations for hundreds of stars from the ground, space-based
photometric surveys have opened the door to homogeneous ensemble measures of
stellar rotation, and therefore age.
Using precise light curves available from the TESS mission, we hope to
infer ages for nearly 100,000 main sequence field stars, including thousands
of planet hosts.
The ages of the planet hosts will allow us to characterize trends in planet
properties over time.

@@

To enable studies of stellar ages and magnetic activity from rotation periods
with TESS, we propose to:
1. Measure accurate rotation periods for every available TESS FFI and
two-minute cadence target.
2. Produce updated gyrochronology relations based on a wider range of field
star ages.
3. Determine the origin of the mysterious rotation period bimodality
discovered with Kepler by tracing the rotation period distribution over the
whole sky, and out to further distances utilizing public Gaia data.
4. Measure the star formation history across the sky using a new Bayesian
age-dating system.
5. Further characterize the intrinsic uncertainty in gyrochronology ages using
comoving pairs identified in the first Gaia data release.

@@

Gyrochronology is by far the most widely applicable, relatively precise
age-dating method available.
Although the phenomenon of magnetic braking has been closely studied since its
discovery in the mid/late 20th century, the manifestations of magnetic fields
and their effects on rotation and activity are complex; the physics
underpinning rotational evolution are far from well-understood.
Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
rapidly than the \citet{skumanich1972} law predicts they should.
\citet{van-saders2016} find that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Once a star reaches a critical Rossby number (the ratio of the rotation period
to the convective overturn time), magnetic braking efficiency effectively
drops to zero.
This means a 5 Gyr star of Solar mass will have the same rotation period as a
6 Gyr and rotation period cannot be used as an age proxy.
The transition occurs at around Solar age for Solar mass stars but later for
less massive stars and earlier for more massive stars.
Although this effect reduces the usefulness of gyrochronology for stars more
massive than the Sun, in practise, for Solar-mass stars, down to M dwarfs with
masses greater than 0.35$M_\odot$, gyrochronology can still be considered an
effective dating method.
This will include a large number of \TESS\ planet candidates.

We intend to use the rotation periods of planet hosting stars to infer the
age-dependence of planet frequency.
Kepler alone has not yet produced a sufficient number of stars with rotation
periods to reliably characterise the time-dependent frequency of exoplanets.
Mann et al. (2016a), Mann et al. (2016b) and Rizzuto et al. (2017) detected
transiting exoplanets in four young open clusters observed by the K2 mission.

>>>>

Angular momentum is carried away though magnetically driven stellar winds,
which slows the star’s rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star’s surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
While rotation periods have previously been measured from starspot-induced
flux modulations for hundreds of stars from the ground, space-based
photometric surveys have opened the door to homogeneous ensemble measures of
stellar rotation, and therefore age.
Using precise light curves available from the TESS mission, we hope to
infer ages for nearly 100,000 main sequence field stars, including a few
hundred planet hosts.
The ages of the planet hosts will allow us to characterize trends in planet
properties over time.

@@

To enable studies of stellar ages and magnetic activity from rotation periods
with TESS, we propose to:
1. Measure accurate rotation periods for every available TESS FFI and
two-minute cadence target.
2. Determine the origin of the mysterious rotation period bimodality
discovered with Kepler by tracing the rotation period distribution over the
whole sky, and out to further distances utilizing public Gaia data.
3. Measure the star formation history across the sky using a new Bayesian
age-dating system.
4. Further characterize the intrinsic uncertainty in gyrochronology ages using
comoving pairs identified in the first Gaia data release.

@@

Gyrochronology is by far the most widely applicable, relatively precise
age-dating method available.
Although the phenomenon of magnetic braking has been closely studied since its
discovery in the mid/late 20th century, the manifestations of magnetic fields
and their effects on rotation and activity are complex; the physics
underpinning rotational evolution are far from well-understood.
Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
rapidly than the \citet{skumanich1972} law predicts they should.
\citet{van-saders2016} find that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Once a star reaches a critical Rossby number (the ratio of the rotation period
to the convective overturn time), magnetic braking efficiency effectively
drops to zero.
This means a 5 Gyr star of Solar mass will have the same rotation period as a
6 Gyr and rotation period cannot be used as an age proxy.
The transition occurs at around Solar age for Solar mass stars but later for
less massive stars and earlier for more massive stars.
Although this effect reduces the usefulness of gyrochronology for stars more
massive than the Sun, in practise, for Solar-mass stars, down to M dwarfs with
masses greater than 0.35$M_\odot$, gyrochronology can still be considered an
effective dating method.

{\bf We will use the rotation periods of TESS field dwarfs to calibrate the
gyrochronology relations using galactic kinematics}.
Many stars in the TESS CTL will also have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate both
galactocentric positions and action angles of the stars, both of which are age
indicators.
Vertical action (angular momentum through the plane of the Milky Way)
increases over time due to orbital heating and has been demonstrated to
increase with age \citep[\eg][]{west2011}.
Galactocentric position can also determine a star's age, since different
stellar populations, thin or thick disk, halo and bulge have different average
ages.
PI Angus is currently using the \kepler\ rotation periods of stars in the
first \Gaia\ data release to calibrate the relations between rotation period
and vertical action dispersion, two tracers of age.
Since this relation is likely vary with both galactic longitude and latitude,
the rotation periods produced from \TESS\ light curves will enable a
comprehensive, all-sky calibration of these relations.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$.

{\bf We will use the rotation periods of planet hosting stars to infer
the age-dependence of planet frequency.}j Kepler alone has not yet produced a
sufficient number of stars with rotation periods to reliably characterise the
time-dependent frequency of exoplanets.
Mann et al. (2016a), Mann et al. (2016b) and Rizzuto et al. (2017) detected
transiting exoplanets in four young open clusters observed by the K2 mission.
